Synthesizing Quotient Lenses

### 2018-03-10 papers/QuotientLenses/main.tex
add eval fixes

We have implemented QREs and the quotient lens synthesis algorithm described
above as an extension to the Boomerang interpreter~\cite{boomerang,quotientlenses}. Thanks to this integration, QREs and quotient lens
synthesis tasks can now be expressed alongside existing Boomerang
terms\bcp{what's the difference between a task and a term (and how do each
  of them relate to ``lenses'' just below)?}, and we
can combine the two kinds of lenses seamlessly in the same application.
Intuition suggests that the increased expressiveness of QREs should make many
lens programming tasks easier, but we would like more quantitative understanding
of that claim.  We thus conducted several experiments that aim to provide
answers to the following questions:

\begin{enumerate}
  \item Does synthesizing quotient lenses from QREs permit an easier development
  process than existing approaches?  \bcp{Needlessly vague: What existing approaches?  Writing
  them by hand?}

  \item Does the increase in expressivity gained from synthesizing from QREs
  permit synthesis of more useful lenses than existing approaches.  \bcp{Is
    this really an interesting empirical question?  Seems obvious, from the
    arguments we made in Section 2 and the intro.}

  \item Does our tool synthesize lenses fast enough to be used as part of a
  standard development process.  \bcp{Should there be some question marks
    around here.}
\end{enumerate}

@@

\item The remaining eight benchmarks were derived from datasets found at 
  data.gov. These datasets include census
  population data, wage comparison data, demographics statistics data and
  crime index data. We chose these datasets since they were stored in XML, CSV
  and RDF formats which are more easily specified using regular expressions than
  plain text or PDF documents.  \BCP{Besides differing formats, why did we
    include these eight new 
    ones?  Are they significantly different (more complex, exercising
    different features of the synthesis algorithm or QRE syntax, ...)?}
\end{enumerate}

@@

To get a sense of how much the new expressiveness that combining QRE lenses with
synthesis saves in terms of programmer effort, we compared the amount of code
that would need to be written for each of the lens benchmarks. We use number of
nodes in the abstract syntax tree as a measure of code size, and calculated
three quantities for each lens synthesis benchmark task. We only compared the
sizes on benchmarks where the canonization functions are nontrivial.
%
\begin{itemize}
  \item[\QRESize{}] the number of AST nodes in the QRE
  specifications (including examples)\saz{CHECK THIS}
  \item[\CanonizerAndSpecSize{}]  the number of
  AST nodes in $W(q_1)$ summed with the number of AST nodes in $\canonizer(q_1)$
  \item[\LensAndSpecSize{}] the number of AST nodes
  in $W(q_1)$ summed with the number of AST nodes in the synthesized QRE lens.
\end{itemize}

@@

Figure~\ref{fig:asts} shows the number of benchmarks that can be expressed using
at most a given number of AST nodes. For a given benchmark, we found \QRESize{}
to be about half as large as \CanonizerAndSpecSize, and about a third as large
as \LensAndSpecSize{}.\BCP{These are really the critical conclusions, and
  they are hard to see, the way the graph is drawn.  I wonder whether we
  could instead present with the individual examples on the X axis (sorted
  by number of AST nodes), with a bar above each one containing three
  regions, colored differently, for LS, CS-LS, and QC-CS...} This suggests
that introducing QREs saves programmers 
significant effort compared to both Optician and basic Boomerang.

@@

The pre-QRE version of Optician was only able to synthesize fully bijective data
transformations.  Synthesizing from QREs loosens that restriction, making it
possible to synthesize data which is bijective when put in canonical
form\bcp{No: We do not synthesize data, nor is the data itself bijective.}. To
determine the increase of expressivity, we analyzed:
\begin{enumerate}
  \item How many data formats in Optician's benchmark suite required modifications
  to be bijective.
  \item How many data formats in Optician's benchmark suite required modifications
  to be bijective when put in canonical form.  \bcp{I don't understand the
    difference between these.}
\end{enumerate}

We found that 10 of Optician's 39 benchmarks required modifications to be
bijective. Of these We found that 4 of Optician's 39 benchmarks required
modifications to be in bijective correspondence when in canonical form.
Furthermore, we found that all eight of the data.gov examples required no
modifications to be in bijective correspondence when in canonical
form.\BCP{Is this surprising?  We could have chosen the examples so that
  this is true.} While we
did not put the data.gov examples into a form suitable for pre-QRE
Optician\bcp{What does that mean?}, we
found that only four of the eight data.gov examples did not require
canonization, so making them suitable for pre-QRE Optician would have required
changes to of four of the data.gov formats. Here again, we see that QREs add
significant expressive power.

@@

\begin{figure}[t]
\centering
\begin{subfigure}[b]{.49\textwidth}
\centering
\includegraphics{generated-graphs/times_opt}
\caption{}
\label{subfig:lenssize}
\end{subfigure}
\begin{subfigure}[b]{.49\textwidth}
\includegraphics{generated-graphs/times_new.eps}
\caption{}
\label{subfig:examplesused}
\end{subfigure}
\caption{Runtimes of Optician and our system.
In (a), we run Optician and our system on Optician's benchmarks.  We find that
there is only a negligible difference incurred by Optometrist's overhead.
In (b), we run Optometrist on its benchmark suite.  We find that it is able
to synthesize all quotient lenses in under 20 seconds, and typically
finished in under 5 seconds. \bcp{The left-hand graph, especially, is hard
  to make anything out of, because the curves are squished against the
  axes and have almost identical shapes.  Not sure what to do about this.
  Also, I'm finding the terms 
  ``Optician'' and ``Optometrist'' really hard to keep straight---in part
  because, in English, they are almost synonyms.}}
\label{fig:times}
\end{figure}

To assess the performance impact of adding QRE support to Optician, we measured
the running time of the following configurations:
\begin{itemize}
  \item[\OpticianRuntime{}] pre-QRE Optician running on its own benchmarks.
  \item[\SystemOnOptician{}] Optician with QREs running on the pre-QRE benchmarks.
  \item[\SystemOnBenchmarks{}] Optician with QREs running on its own
  benchmarks. \BCP{What is the difference between QO and QQ??  The eight
    additional benchmarks?  Why not measure just these?  Seems like this
    graphj repeats some information from the one to the left, obscuring the
    new information.}
\end{itemize}

@@

Also notable is one of the data.gov examples, which converts demographic
statistics by zip code from xml form into rdf form.\bcp{It's a bit hidden in the figure...} To validate this performance
was not due to our overhead, we modified this format to be bijective, so it
could be input into Optician. Optician took over 15 seconds synthesize this
modified lens. We believe this conversion took longer because the input formats
were exceptionally large, requiring conversions between over 40 fields.

>>>>

We have implemented QREs and the quotient lens synthesis algorithm described
above as an extension to the Boomerang
interpreter~\cite{boomerang,quotientlenses}. We have added QREs and QRE lens
synthesis tasks to Boomerang. Furthermore, these synthesis tasks evaluate to
Boomerang lens values, so Boomerang treats synthesized lenses the same as
hand-written ones. Intuition suggests that the increased expressiveness
of QREs should make many lens programming tasks easier, but we would like more
quantitative understanding of that claim. We thus conducted several experiments
that aim to provide answers to the following questions:

\begin{enumerate}
  \item Does synthesizing QRE lenses from QREs permit an easier development
  process than manually writing canonizers, and synthesizing the lenses between
  their canonized forms?  Does synthesizing QRE lenses from QREs permit an
  easier development process than writing the lenses by hand?

  \item Does the increase in expressivity gained from synthesizing from QREs
  permit synthesis of more useful lenses than existing approaches.  \bcp{Is
    this really an interesting empirical question?  Seems obvious, from the
    arguments we made in Section 2 and the intro.}\afm{I'm happy to remove it}

  \item Does our tool synthesize lenses fast enough to be used as part of a
  standard development process?
\end{enumerate}

@@

\item The remaining eight benchmarks were derived from datasets found at 
  data.gov. These datasets include census
  population data, wage comparison data, demographics statistics data and
  crime index data. We chose these datasets since they were stored in XML, CSV
  and RDF formats which are more easily specified using regular expressions than
  plain text or PDF documents.  \BCP{Besides differing formats, why did we
    include these eight new 
    ones?  Are they significantly different (more complex, exercising
    different features of the synthesis algorithm or QRE syntax,
    ...)?}\afm{Solomon, could you fill this out?}
\end{enumerate}

@@

To get a sense of how much the new expressiveness that combining QRE lenses with
synthesis saves in terms of programmer effort, we compared the amount of code
that would need to be written for each of the lens benchmarks. We use number of
nodes in the abstract syntax tree as a measure of code size, and calculated
three quantities for each lens synthesis benchmark task. We only compared the
sizes on benchmarks where the canonization functions are nontrivial.
%
\begin{itemize}
  \item[\QRESize{}] the number of AST nodes in the QRE
  specifications (including examples)\saz{CHECK THIS}\afm{no but i'll add it in}
  \item[\CanonizerAndSpecSize{}]  the number of
  AST nodes in $W(q_1)$ summed with the number of AST nodes in $\canonizer(q_1)$
  \item[\LensAndSpecSize{}] the number of AST nodes
  in $W(q_1)$ summed with the number of AST nodes in the synthesized QRE lens.
\end{itemize}

@@

Figure~\ref{fig:asts} shows the number of benchmarks that can be expressed using
at most a given number of AST nodes. For a given benchmark, we found \QRESize{}
to be about half as large as \CanonizerAndSpecSize, and about a third as large
as \LensAndSpecSize{}.\BCP{These are really the critical conclusions, and
  they are hard to see, the way the graph is drawn.  I wonder whether we
  could instead present with the individual examples on the X axis (sorted
  by number of AST nodes), with a bar above each one containing three
  regions, colored differently, for LS, CS-LS, and QC-CS...}\afm{will do} This suggests
that introducing QREs saves programmers 
significant effort compared to both Optician and basic Boomerang.

@@

The pre-QRE version of Optician was only able to synthesize fully bijective data
transformations. Synthesizing from QREs loosens that restriction, making it
possible to synthesize data transformations involving bijections with
canonization functions. Because Optician could previously only synthesize
bijective lenses, some of the data formats in the example suite were modified to
include additional fields containing information should should be canonized from
the other format. In total, 12 of Optician's original benchmarks and 3 of the
data.gov benchmarks would require such additional fields when input to Optician
without QREs.

@@

\begin{figure}[t]
\centering
\begin{subfigure}[b]{.49\textwidth}
\centering
\includegraphics{generated-graphs/times_opt}
\caption{}
\label{subfig:lenssize}
\end{subfigure}
\begin{subfigure}[b]{.49\textwidth}
\includegraphics{generated-graphs/times_new.eps}
\caption{}
\label{subfig:examplesused}
\end{subfigure}
\caption{Runtimes of Optician and our system. In (a), we run Optician and our
  system on Optician's benchmarks. We find that there is only a negligible
  performance overhead incurred by using QREs. In (b), we run Optician with QREs
  on its benchmark suite. We find that it is able to synthesize all quotient
  lenses in under 20 seconds, and typically finished in under 5 seconds.
  \bcp{The left-hand graph, especially, is hard to make anything out of, because
    the curves are squished against the axes and have almost identical shapes.
    }\afm{will fix}}
\label{fig:times}
\end{figure}

To assess the performance impact of adding QRE support to Optician, we measured
the running time of the following configurations:
\begin{itemize}
  \item[\OpticianRuntime{}] pre-QRE Optician running on its own benchmarks.
  \item[\SystemOnOptician{}] Optician with QREs running on the pre-QRE benchmarks.
  \item[\SystemOnBenchmarks{}] Optician with QREs running on its own
  benchmarks. \BCP{What is the difference between QO and QQ??  The eight
    additional benchmarks?  Why not measure just these?  Seems like this
    graphj repeats some information from the one to the left, obscuring the
    new information.}\afm{will do}
\end{itemize}

@@

Also notable is one of the data.gov examples, which converts demographic
statistics by zip code from xml form into rdf form. Synthesizing this
transformation certainly took the longest, nearly fifteen seconds, where the
next slowest took hardly over five. To validate this performance was not due to
our overhead, we modified these format to be bijective, so the synthesis task
could be input into Optician. Optician took over 15 seconds synthesize this
modified lens. We believe this conversion took longer because the input formats
were exceptionally large, requiring conversions between over 40 fields, not due
to an overhead from synthesizing from QREs.
